---
title: "AMR_rf_models_GIT"
author: "John Berini"
date: "2026-02-14"
output: html_document
---

bring in our data and load the packages
```{r setup}
#packages
library(MixRF)
library(randomForest)
library(caret)
library(dplyr)#top_n
library(tidyverse)#rownames_to_column
library(Hmisc) #rcorr
library(ranger)
library(car)#vif function
library(sp)
library(sf)
library(gstat)
library(ggplot2)


amr<-read.csv("data/AMR_2024.12.09.csv", header=T) #need to create a file that keeps the number of lakes at 39 and does not reduce. 
covs<-read.csv("data/rf_dat.csv", header=T) #create a new covs file that has all 39 lakes found in the amr dataframe

vi<-st_read("gis/vi_outline.shp")#outline of Vancouver Island for clipping spatial data later


#regression kriging only works with continuous data, so we will remove those covariates that are non continuous
covs <- covs[, !names(covs) %in% c("cov_patt", "crwn_clsre")] #these are ordinal


```


first we need to subset and aggregate the original AMR data, as each lake has three different data points. 

```{r}
lamr<-amr
colnames(lamr)[which(names(lamr) == "Lake")] <- "lake_code"

var_list<-names(lamr) %in% c("Samples", "lake_code", "Lake_AMR", "co_occur")
lamr<-lamr[var_list]
```

aggregate data to get the means and remove those that have ZEROS remaining

```{r}
#aggregate the data based on lake code
lamr_agg<-as.data.frame(aggregate(lamr$Lake_AMR, list(lamr$lake_code), FUN=mean))

#rename columns
names(lamr_agg)<-c("lake_code", "lake_amr")

#remove the zeroes
lamr_agg<-lamr_agg[lamr_agg$lake_amr>0,] #remove lake_amr values equal to 0
lamr_agg <- lamr_agg[complete.cases(lamr_agg), ] #remove NAs

```

Make sure that the lakes are the same in the COVS dataframe and the AMR dataframe and vice versa

```{r}
unique(lamr_agg$lake_code[!(lamr_agg$lake_code%in%covs$lake_code)]) #list of amr$lake_codes not in covs

lamr_agg<-lamr_agg[lamr_agg$lake_code %in% covs$lake_code,] #remove any lake_codes not in covs$lake_codes

unique(lamr_agg$lake_code[!(lamr_agg$lake_code%in%covs$lake_code)]) #recheck --> should have removed all lake_code in lamr not in covs

unique(covs$lake_code[!(covs$lake_code%in%lamr_agg$lake_code)]) #list of amr$lake_codes not in covs

covs<-covs[covs$lake_code %in% lamr_agg$lake_code,] #remove any ar$lake_codes not in covs$lake_codes

unique(covs$lake_code[!(covs$lake_code%in%lamr_agg$lake_code)])
```

now we merge the datasets

```{r}
#merge
lamr_merge<-merge(x = lamr_agg, y = covs, by = "lake_code")

#remove columns from merged data frame that are not useful. 
lake_code<-lamr_merge$lake_code
lake_name<-lamr_merge$lake_name
longitude<-lamr_merge$longitude
latitude<-lamr_merge$latitude

#subset lamr_merge to remove those columns not wanted or needed in random forest and model selection
lamr_merge <- lamr_merge[, -which(names(lamr_merge) %in% c("lake_code", "X", "lake_name", "lake_name.1", "latitude", "longitude"))]

lamr_rf_dat<-lamr_merge
```

First we need to clean up our covariate data 

```{r}
length(lamr_rf_dat)
#select only those covariates that are numeric, omitting names, etc.
lamr_rf_dat<- lamr_rf_dat %>% select_if(is.numeric)
length(lamr_rf_dat)

#remove any columns that consist of all zeros
lamr_rf_dat <- lamr_rf_dat %>% select_if(~ sum(.x, na.rm = TRUE) > 0)
length(lamr_rf_dat)

#remove covariates with a variance of 0
lamr_rf_dat <- lamr_rf_dat %>% select_if(~ sd(.x) != 0)
length(lamr_rf_dat)

#create a separate object for lake_amr and all other data for the boruta algorithm
lake_amr<-lamr_rf_dat$lake_amr
rf_dat <- lamr_rf_dat[, -which(names(lamr_rf_dat) %in% c("lake_amr"))]
```

Below we are using the ranger function implement random forest for variable selection. However, because random forest has randomness as an inherent part of the process, different runs of the same process can lead to a different set of variables. To alleviate this, we are looping over 100 iterations of the ranger process and pulling out those covariates that have the highest mean importance. 

```{r}
# Sub-function: remove collinear predictors based on importance from ranger
remove_collinear_by_importance <- function(data, response, cutoff = 0.7, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  full_data <- data.frame(response = response, data)

  rf_model <- ranger(response ~ ., data = full_data, importance = "permutation")
  importance_scores <- rf_model$variable.importance

  data_num <- data[, sapply(data, is.numeric)]
  cor_matrix <- cor(data_num, use = "pairwise.complete.obs")

  to_remove <- c()

  repeat {
    high_corr_pairs <- which(abs(cor_matrix) > cutoff & upper.tri(cor_matrix), arr.ind = TRUE)
    if (nrow(high_corr_pairs) == 0) break

    var1 <- colnames(cor_matrix)[high_corr_pairs[1, 1]]
    var2 <- colnames(cor_matrix)[high_corr_pairs[1, 2]]

    imp1 <- importance_scores[var1]; if (is.null(imp1)) imp1 <- -Inf
    imp2 <- importance_scores[var2]; if (is.null(imp2)) imp2 <- -Inf

    remove_var <- if (imp1 >= imp2) var2 else var1
    to_remove <- c(to_remove, remove_var)

    cor_matrix <- cor_matrix[!(rownames(cor_matrix) == remove_var), !(colnames(cor_matrix) == remove_var)]
    importance_scores <- importance_scores[!(names(importance_scores) == remove_var)]
  }

  cleaned_data <- data[, !(names(data) %in% to_remove)]
  return(cleaned_data)
}

# Wrapper: repeat the above function n_iter times and retain stable predictors
remove_collinear_by_stability <- function(data, response, cutoff = 0.7, n_iter = 1000, threshold = 0.9, seed = 42) {
  set.seed(seed)
  seeds <- sample(1e6, n_iter)

  retained_list <- vector("list", n_iter)
  pb <- txtProgressBar(min = 0, max = n_iter, style = 3)

  for (i in seq_len(n_iter)) {
    cleaned <- remove_collinear_by_importance(data, response, cutoff, seed = seeds[i])
    retained_list[[i]] <- names(cleaned)
    setTxtProgressBar(pb, i)
  }

  close(pb)

  # Tabulate how often each variable was retained
  all_vars <- unlist(retained_list)
  var_freq <- table(all_vars) / n_iter

  # Keep only variables that were retained in at least 'threshold' proportion of runs
  stable_vars <- names(var_freq[var_freq >= threshold])

  cat("\nFinal retained variables (stability ≥", threshold, "):\n")
  print(stable_vars)

  # Return stable variable subset of the original data
  return(list(
    data = data[, stable_vars, drop = FALSE],
    retained_vars = stable_vars,
    retention_frequency = var_freq
  ))
}

#run custom function above
noncollinear_covs <- remove_collinear_by_stability(rf_dat, lake_amr, cutoff = 0.7, n_iter = 1000, threshold = 0.9, seed = 123)

#subset rf_dat to create a new dataset that contains only those important covariates 
clean_rf_dat <- rf_dat[, noncollinear_covs$retained_vars]

#=====================================================================================
#above we used ranger to establish covariate importance and used that to determine which covariates to keep. Here, we will use ranger to evaluate which covaraites are most important in AMR prediction.
#Looping ranger for stable variable selection:
n_runs <- 1000
top_k <- 20
top_vars_list <- list()

for (i in 1:n_runs) {
 set.seed(i)  # set seed for reproducibility

 rf <- ranger(
   formula = lake_amr ~ .,
   data = clean_rf_dat,
   importance = "permutation",  # better for variable selection
   num.trees = 1000
 )

 # Get variable importance and extract top variables
 importance_df <- data.frame(variable = names(rf$variable.importance),
                             importance = rf$variable.importance) %>%
   arrange(desc(importance)) %>%
   slice_head(n = top_k)

 top_vars_list[[i]] <- importance_df$variable
}

# Combine all selected variables into a single vector
all_top_vars <- unlist(top_vars_list)

# Count how often each variable was selected
variable_freq <- sort(table(all_top_vars), decreasing = TRUE)

# View the most stable variables
head(variable_freq, 20)

# create list of important covariates
important_covariates<-names(head(variable_freq, 20))

# extract data for important covariates from our rf_dat
important_predictors <- rf_dat[, important_covariates]

```

Now we can use the step function to find our best fitting model

```{r}
#Use step function to find best fitting model from ranger function
  full_AMR_model_rngr <- lm(lake_amr~.,data=important_predictors)
  #use the step function to perform both forward and backward selection
  step_AMR_model_rngr <- step(full_AMR_model_rngr, direction = "both") #performs both backward and forward selection 
  summary_step_AMR_rngr<-summary(step_AMR_model_rngr) #creates an object of the summary statistics
  summary(full_AMR_model_rngr)
  
ranger_fit<-step_AMR_model_rngr
summary(ranger_fit)
```


Now that we have our best fitting model,we can begin the process of regression kriging.

First we'll install some packages. 

```{r}
library(gstat)
library(terra)
library(sp)
library(sf)
library(raster)
```

Now we need to read in our spatial data, based in our best fitting model. 
Note that data are too large to be hosted on GitHUB and are hosted on a local network folder. 

```{r}
#best fitting model based on ranger and step functions
summary(ranger_fit)

#make sure to attached the courses drive @ Carleton
#landscape covariates
d2_lake<-rast("tiff/d2_lake.tif")

d2_ancient<-rast("/tiff/d2_ancient.tif")

spp3_cvr<-rast("tiff/spp3_cvr.tif")

#climatic covariates
mean_NFFD09<-rast("tiff/BC/800m/Normal_1991_2020/NFFD09.tif")

```

check the raster coordinate systems and make sure they are all the same extent and resolution

```{r}
#coordinate systems
crs(d2_lake)
crs(d2_ancient)
crs(spp3_cvr)
crs(mean_NFFD09)

#check resolution and resample if necessary. Resampling to the coarsest raster to save time/computational resources
res(d2_lake)
d2_lake <- resample(d2_lake, mean_NFFD09, method = "bilinear")

res(d2_ancient)
d2_ancient <- resample(d2_ancient, mean_NFFD09, method = "bilinear")

res(spp3_cvr)
spp3_cvr <- resample(spp3_cvr, mean_NFFD09, method = "bilinear")

res(mean_NFFD09)

#now we just need to check our extents
ext(d2_lake)
ext(d2_ancient)
ext(spp3_cvr)
ext(mean_NFFD09)

#check names
names(d2_lake)
names(d2_ancient)
names(spp3_cvr)
names(mean_NFFD09)

```

Calculate residuals and compute a variogram

```{r}
#estimate our residuals for our best fitting model
rf_dat$residuals<-residuals(ranger_fit)

#replace locational and identification infomration for each lake
rf_dat$lake_code<-lake_code
rf_dat$lake_name<-lake_name

# Create a SpatialPointsDataFrame from rf_dat
coordinates(rf_dat) <- ~longitude + latitude

# Assign the CRS from your raster (converted to PROJ.4 format)
proj4string(rf_dat) <- crs(d2_lake, proj = TRUE)

#subset rf_dat to contain only important_covariates
rf_dat <- rf_dat[, c("d2_lake", "d2_ancient", "spp3_cvr", "mean_NFFD09", "residuals")]

# Create experimental variogram of residuals
vgram <- variogram(residuals ~ 1, rf_dat)

# Fit variogram model
vgram_model <- fit.variogram(vgram, model = vgm("Sph"))  # or try "Exp", "Gau", etc.
plot(vgram, vgram_model)

```

create an empty grid based on my current rasters

```{r}
# --- Create an empty grid that matches your covariate raster ---

# Get the resolution/extent/CRS from your base raster
raster_ext <- ext(d2_lake)
raster_res <- res(d2_lake)
raster_crs <- crs(d2_lake)

# Make an empty raster template
empty_grid <- rast(raster_ext, resolution = raster_res, crs = raster_crs)

# --- Convert raster grid to prediction locations for kriging ---

# Extract coordinates for every cell
coords <- xyFromCell(empty_grid, 1:ncell(empty_grid))

# Build a SpatialPixelsDataFrame (preferred input for gstat::krige)
pred_df <- data.frame(x = coords[,1], y = coords[,2])
coordinates(pred_df) <- ~x+y
gridded(pred_df) <- TRUE
proj4string(pred_df) <- CRS(as.character(raster_crs))

```

Now that we have an empty grid within with to predict our kriged values, we can perform kriging of our residuals

```{r}
#1.	Stack your predictor rasters:
preds_stack <- c(d2_lake, d2_ancient, spp3_cvr, mean_NFFD09)
names(preds_stack) <- c("d2_lake", "d2_ancient", "spp3_cvr", "mean_NFFD09")

#2.	Predict regression surface across the raster stack:
reg_preds_raster <- predict(preds_stack, ranger_fit)  # terra handles rasters natively

#3.	Create a prediction grid for residual kriging:
# Use the same raster as template
empty_grid <- preds_stack[[1]]  # pick any layer as template

# Convert to SpatialPixelsDataFrame for gstat
coords <- xyFromCell(empty_grid, 1:ncell(empty_grid))
pred_df <- data.frame(x = coords[,1], y = coords[,2])
coordinates(pred_df) <- ~x+y
gridded(pred_df) <- TRUE
proj4string(pred_df) <- CRS(as.character(crs(empty_grid)))

#4.	Kriging residuals on that grid:
# Compute residuals
residuals <- lake_amr - predict(ranger_fit, newdata = rf_dat)

# Add residuals as a new column in your SpatialPointsDataFrame
rf_dat$resid <- residuals

# Now rf_dat is ready for kriging
res_kriged <- krige(resid ~ 1, rf_dat, pred_df, model = vgram_model,
                    nmax = Inf, maxdist = Inf)

#5.	Combine regression and kriging surfaces:
# Convert reg_preds_raster to vector to align with kriged
reg_vals <- values(reg_preds_raster)

# Regression + residuals
combined_vals <- reg_vals + res_kriged$var1.pred

# Put back into a raster
reg_kriged_raster <- empty_grid
values(reg_kriged_raster) <- combined_vals

reg_kriged_raster <- mask(reg_kriged_raster, vi)
plot(reg_kriged_raster)

# Plot
plot(reg_kriged_raster,
     xlim = c(-129, -123),  # xmin, xmax
     ylim = c(48, 51),    # ymin, ymax
     main = "Regression-Kriging Raster (Zoomed In)")
```

Now that we've completed our kriging of the regression residuals and used our best fitting model to predict AMR values, we need to export these data as a raster. 

```{r}
library(terra)

# Define output path
out_file <- "R_out/gis/regression_kriging_result.tif"

# Write the raster
writeRaster(reg_kriged_raster,
            filename = out_file,
            overwrite = TRUE)   # overwrite if needed
```

Now that we have out best fitting model and it is converted to a raster, we want to perform spatial hold-one out (a.k.a.- leave one out cross validation). 

```{r}
###############################################################################
# Spatial Leave-One-Out Cross-Validation (LOOCV) for Regression-Kriging
###############################################################################

# -----------------------------
# 1. Prepare
# -----------------------------

rf_dat@data$lake_amr <- lake_amr

# Name of response column
response_col <- "lake_amr"

# Ensure response column is in @data
if (!response_col %in% names(rf_dat@data)) {
  rf_dat@data[[response_col]] <- rf_dat[[response_col]]
}

# Predictor columns are all columns in @data except the response
predictor_cols <- setdiff(names(rf_dat@data), response_col)

# Check there is at least one predictor
if (length(predictor_cols) == 0) {
  stop("No predictor columns found! Cannot fit random forest.")
}

n_points <- nrow(rf_dat)
predictions <- numeric(n_points)  # store LOOCV predictions


# -----------------------------
# 2. Spatial LOOCV loop
# -----------------------------
for (i in 1:n_points) {
  
  cat("Processing point", i, "of", n_points, "\n")
  
  # 2a. Split data
  train_data <- rf_dat[-i, ]
  test_data  <- rf_dat[i, , drop = FALSE]
  
  # 2b. Prepare training and test data frames for RF
  rf_train_df <- data.frame(train_data@data[, predictor_cols, drop = FALSE],
                            lake_amr = train_data@data[, response_col])
  rf_test_df <- data.frame(test_data@data[, predictor_cols, drop = FALSE])
  
  # 2c. Fit RF on training data
  rf_fit <- ranger(lake_amr ~ ., data = rf_train_df)
  
  # 2d. Compute residuals on training data
  train_resid <- train_data@data[, response_col] - predict(rf_fit, data = rf_train_df)$predictions
  train_data$resid <- train_resid
  
  # 2e. Fit variogram on residuals
  vgm_model <- variogram(resid ~ 1, train_data)
  vgm_fit   <- fit.variogram(vgm_model, model = vgm("Sph"))
  
  # 2f. Krige residual to held-out point
  res_kriged <- krige(resid ~ 1, train_data, test_data, model = vgm_fit)
  
  # 2g. Predict regression value for held-out point
  rf_pred <- predict(rf_fit, data = rf_test_df)$predictions
  
  # 2h. Combine regression + kriged residual
  predictions[i] <- rf_pred + res_kriged$var1.pred
}

# -----------------------------
# 3. Evaluate performance
# -----------------------------
observed <- rf_dat@data[, response_col]
rmse <- sqrt(mean((predictions - observed)^2))
correlation <- cor(predictions, observed)

cat("LOOCV RMSE:", rmse, "\n")
cat("LOOCV correlation:", correlation, "\n")

# -----------------------------
# 4. Store results
# -----------------------------
rf_dat$LOOCV_pred <- predictions
```

Now we can make some diagnostic plots to show how well out model is working. 

```{r}
###############################################################################
# Spatial LOOCV Diagnostic Plots
###############################################################################

# Extract observed and predicted values
observed <- rf_dat@data$lake_amr
predicted <- rf_dat$LOOCV_pred
residuals <- predicted - observed

# 1. Observed vs Predicted
p1 <- ggplot(data = data.frame(observed, predicted), aes(x = observed, y = predicted)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1.2) +
  labs(x = "Observed lake_amr", y = "Predicted (LOOCV)",
       title = "a)") +
  theme_minimal()
print(p1)

# Add RMSE and correlation as text
rmse <- sqrt(mean(residuals^2))
correlation <- cor(predicted, observed)
p1 <- p1 + annotate("text", x = min(observed), y = max(predicted),
                    label = paste0("RMSE = ", round(rmse,2), "\nR = ", round(correlation,2)),
                    hjust = 0, vjust = 1, size = 4)

# 2. Residuals vs Predicted
p2 <- ggplot(data = data.frame(predicted, residuals), aes(x = predicted, y = residuals)) +
  geom_point(color = "darkgreen", alpha = 0.7) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", size = 1) +
  labs(x = "Predicted lake_amr", y = "Residuals",
       title = "b)") +
  theme_minimal()
print(p2)

# 3. Histogram of residuals
p3 <- ggplot(data = data.frame(residuals), aes(x = residuals)) +
  geom_histogram(bins = 15, fill = "lightgray", color = "black") +
  labs(x = "Residuals", y = "Count",
       title = "c)") +
  theme_minimal()
print(p3)

# 4. Spatial plot of residuals
# Extract coordinates and residuals
predicted <- rf_dat$LOOCV_pred
observed <- rf_dat@data$lake_amr
residuals <- predicted - observed

# Combine coordinates + residuals into a single data frame
resid_df <- data.frame(
  x = rf_dat@coords[,1],
  y = rf_dat@coords[,2],
  residuals = residuals
)

# Check if vi is sf or sp
if (inherits(vi, "sf")) {
  vi_sf <- vi
} else if (inherits(vi, "Spatial")) {
  vi_sf <- st_as_sf(vi)  # convert sp to sf
} else {
  stop("vi must be an sf or Spatial object")
}

# Plot spatial residuals with Vancouver Island outline
p4 <- ggplot() +
  geom_point(data = resid_df, aes(x = x, y = y, color = residuals), size = 2) +
  geom_sf(data = vi_sf, fill = NA, color = "black", size = 0.8) +  # outline
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(color = "Residuals", title = "d)") +
  theme_minimal() +
  coord_sf()  # ensures correct aspect ratio for spatial data

print(p4)

library(patchwork)

# Arrange plots
(p1 | p2) / (p3 | p4)

```

