---
title: "AMR_RF_Hypothesis_Models"
author: "John Berini"
date: "2026-02-21"
output: html_document
editor_options: 
  chunk_output_type: console
---

bring in our data and load the packages
```{r setup}
#packages
library(caret)
library(dplyr)#top_n
library(tidyverse)#rownames_to_column
library(Hmisc) #rcorr
library(car)#vif function
library(sp)
library(sf)
library(gstat)
library(ggplot2)


amr<-read.csv("data/AMR_2024.12.09.csv", header=T) #need to create a file that keeps the number of lakes at 39 and does not reduce. 

covs<-read.csv("data/rf_dat.csv", header=T) #create a new covs file that has all 39 lakes found in the amr dataframe

vi_vect<-st_read("/Volumes/jberini_SSD/spat_dat/gis/acoustic monitor locations/vi_outline.shp")
#vi<-st_read("gis/vi_outline.shp")

#regression kriging only works with continuous data, so we will remove those covariates that are non continuous
covs <- covs[, !names(covs) %in% c("cov_patt", "crwn_clsre")] #these are ordinal

```

subset data to include only those covariates that we are interested in:
watershed
underlying bed rock
land-use variables
  - % forest Loss
  - distance fo nearest road
  - distance to aquaculture
Environmental Covariates
  - air temperature (mean?)
  - lake depth (fetch size) - lakemorpho package
Distance to nearest lake
  - some way to include slope/topography
  - 
```{r}
covs_subset <- covs[, c(
  "lake_name",
  "lake_code",
  "latitude",
  "longitude",
  "watershed",
  "bedrock",
  "mean_TREE_COVER",
  "mean_SOIL_NUTRI",
  "mean_d2_paved",
  "mean_d2_ogdc",
  "mean_rd_dnse_loose",
  "mean_rd_dnse_ogdc",
  "mean_rd_dnse_paved",
  "d2_aquaculture",
  "mean_Tmax_sm",
  "max_depth_est",
  "mean_d2_lake",
  "mean_DEM_vi",
  "mean_Band_1",
  "mean_spp1_pc",
  "mean_NFFD09",
  "mean_spp2_pc",
  "mean_basal_area",
  "mean_SPECIES__3"
)]

```

```{r}
lamr<-amr
colnames(lamr)[which(names(lamr) == "Lake")] <- "lake_code"

var_list<-names(lamr) %in% c("Samples", "lake_code", "Lake_AMR", "co_occur")
lamr<-lamr[var_list]
```

aggregate data to get the means and remove those that have ZEROS remaining

```{r}
#aggregate the data based on lake code
lamr_agg<-as.data.frame(aggregate(lamr$Lake_AMR, list(lamr$lake_code), FUN=mean))

#rename columns
names(lamr_agg)<-c("lake_code", "lake_amr")

#remove the zeroes
lamr_agg<-lamr_agg[lamr_agg$lake_amr>0,] #remove lake_amr values equal to 0
lamr_agg <- lamr_agg[complete.cases(lamr_agg), ] #remove NAs

```

now we merge the datasets

```{r}
#merge
lamr_merge<-merge(x = lamr_agg, y = covs_subset, by = "lake_code")

#remove columns from merged data frame that are not useful. 
lake_code<-lamr_merge$lake_code
lake_name<-lamr_merge$lake_name
longitude<-lamr_merge$longitude
latitude<-lamr_merge$latitude
lake_amr<-lamr_merge$lake_amr

#subset lamr_merge to remove those columns not wanted or needed in random forest and model selection
lamr_merge <- lamr_merge[, -which(names(lamr_merge) %in% c("lake_code", "X", "lake_name", "lake_name.1", "latitude", "longitude"))]

lamr_rf_dat<-lamr_merge
```

subset lamr_rf_dat

```{r}
lamr_subset <- lamr_rf_dat[, c(
  "watershed",
  "bedrock",
  "mean_TREE_COVER",
  "mean_SOIL_NUTRI",
  "mean_d2_paved",
  "mean_d2_ogdc",
  "mean_rd_dnse_loose",
  "mean_rd_dnse_ogdc",
  "mean_rd_dnse_paved",
  "d2_aquaculture",
  "mean_Tmax_sm",
  "max_depth_est",
  "mean_d2_lake",
  "mean_DEM_vi",
  "mean_Band_1",
  "mean_spp1_pc",
  "mean_NFFD09",
  "mean_spp2_pc",
  "mean_basal_area",
  "mean_SPECIES__3"
)]


lamr_subset$lake_code<-lake_code
lamr_subset$latitude<-latitude
lamr_subset$longitude<-longitude
lamr_subset$lake_amr<-lake_amr

```


perform model selection
```{r}
library(lmerTest)
# 1. Define your candidate column names
candidates <- c(  "mean_d2_paved",
                  "mean_d2_ogdc",
                  "mean_rd_dnse_loose",
                  "mean_rd_dnse_ogdc",
                  "mean_rd_dnse_paved",
                  "d2_aquaculture",
                  "mean_Tmax_sm",
                  "mean_d2_lake",
                  "mean_SOIL_NUTRI",
                  "mean_TREE_COVER",
                  "mean_Band_1",
                  "mean_spp1_pc",
                  "bedrock",
                  "mean_spp2_pc",
                  "mean_NFFD09",
                  "max_depth_est",
                  "mean_DEM_vi", 
                  "mean_basal_area",
                  "mean_SPECIES__3"
)

# 2. Setup results table
results <- data.frame(Variable = character(), 
                      P_Value = numeric(), 
                      Estimate = numeric(), 
                      Warnings = character(),
                      stringsAsFactors = FALSE)

# 3. The Fixed Loop
for (col_name in candidates) {
  
  # Construct the formula string safely
  form_str <- paste0("lake_amr ~ scale(", col_name, ") + (1 | bedrock)")
  form <- as.formula(form_str)
  
  # Run the model
  # tryCatch allows us to capture warnings like "singular fit"
  try({
    m <- lmer(form, data = lamr_subset)
    
    # Check for singular fit (common with small n)
    is_singular <- isSingular(m)
    
    # Extract stats
    # Fixed effects table: Row 1 is Intercept, Row 2 is your variable
    summ <- summary(m)$coefficients
    p_val <- summ[2, "Pr(>|t|)"]
    est <- summ[2, "Estimate"]
    
    # Save results
    results <- rbind(results, data.frame(
      Variable = col_name, 
      P_Value = p_val, 
      Estimate = est,
      Warnings = if(is_singular) "Singular Fit" else "None"
    ))
  }, silent = TRUE)
}

# 4. View your "Shortlist"
results <- results[order(results$P_Value), ]
print(results)


# 1. Define the Null Model (No predictors, just the random effect)
null_model <- lmer(lake_amr ~ (1 | bedrock), data = lamr_subset, REML = FALSE)

# 2. Re-run your Full Model with REML = FALSE 
# (Required for comparing models with different fixed effects)
model1 <- lmer(lake_amr ~ scale(mean_d2_lake) + 
                         scale(mean_SPECIES__3) + 
                         scale(d2_aquaculture) +
                         scale(mean_Band_1) + #distance to old growth forest
                         (1|bedrock),
              data = lamr_subset)
summary(model1)

anova(null_model, model1)

```

Use best fitting model in a mixed-effets regression kriging format. 

Now that we have our best fitting model,we can begin the process of regression kriging.

First we'll install some packages. 

```{r}
library(gstat)
library(terra)
library(sp)
library(sf)
library(raster)
```

Now we need to read in our spatial data, based in our best fitting model. 

```{r}
#best fitting model based on ranger and step functions
summary(model1)

#make sure to attached the courses drive @ Carleton
#landscape covariates
d2_lake<-rast("/Volumes/COURSES/biol394-13-f23/StuWork/johnsonk4/raster/wgs84UTMz10/tiff/d2_lake.tif")
#d2_lake<-rast("tiff/d2_lake.tif")

d2_ancient<-rast("/Volumes/COURSES/biol394-13-f23/StuWork/johnsonk4/raster/wgs84UTMz10/tiff/d2_ancient.tif")
#d2_ancient<-rast("/tiff/d2_ancient.tif")

spp3_cvr<-rast("/Volumes/COURSES/biol394-13-f23/StuWork/johnsonk4/raster/wgs84UTMz10/tiff/spp3_cvr.tif")
#spp3_cvr<-rast("tiff/spp3_cvr.tif")

#distance to aquaculture
d2_aqua<-rast("data/d2_aquaculture.tif")

#watershed
bedrock<-st_read("../../gis/BC_digital_geology_shp/BC_bedrock_ll83.shp")
# fix invalid polygons
bedrock <- st_make_valid(bedrock)
# remove any empty geometries created by st_make_valid
bedrock <- bedrock[!st_is_empty(bedrock), ]


```

check the raster coordinate systems and make sure they are all the same extent and resolution

```{r}
#coordinate systems
crs(d2_lake)
crs(d2_ancient)
crs(spp3_cvr)
crs(d2_aqua)
crs(bedrock)

st_crs(bedrock)
st_crs(vi_vect)

#check resolution and resample if necessary. Resampling to the coarsest raster to save time/computational resources
res(d2_lake)

res(d2_ancient)
d2_ancient <- resample(d2_ancient, d2_lake, method = "bilinear")

res(spp3_cvr)
spp3_cvr <- resample(spp3_cvr, d2_lake, method = "bilinear")

res(d2_aqua)
d2_aqua <- resample(d2_aqua, d2_lake, method = "bilinear")

# Now crop to Vancouver Island
spp3_cvr   <- crop(spp3_cvr, vi_vect)
d2_lake    <- crop(d2_lake, vi_vect)
d2_ancient <- crop(d2_ancient, vi_vect)
d2_aqua    <- crop(d2_aqua, vi_vect)

# Final alignment step (guarantees identical grids)
d2_ancient <- resample(d2_ancient, d2_lake)
spp3_cvr   <- resample(spp3_cvr, d2_lake)
d2_aqua    <- resample(d2_aqua, d2_lake)

#now we just need to check our extents
ext(d2_lake)
ext(d2_ancient)
ext(spp3_cvr)
ext(d2_aqua)
ext(bedrock)

#check names - if incorrect, assign appropriate name
names(d2_lake)
names(d2_ancient)
# Rename it
names(d2_ancient) <- "d2_ancient"

names(spp3_cvr)
# Rename it
names(spp3_cvr) <- "spp3_cvr"


names(d2_aqua)
# Rename it
names(d2_aqua) <- "d2_aqua"

#check names in lamr_subset
names(lamr_subset)

#rename data
names(lamr_subset)[names(lamr_subset) == "mean_d2_lake"] <- "d2_lake"
names(lamr_subset)[names(lamr_subset) == "mean_Band_1"] <- "d2_ancient"
names(lamr_subset)[names(lamr_subset) == "mean_SPECIES__3"] <- "spp3_cvr"
names(lamr_subset)[names(lamr_subset) == "d2_aquaculture"] <- "d2_aqua"

```

Calculate residuals and compute a variogram

```{r}
#estimate our residuals for our best fitting model
lamr_subset$residuals<-residuals(model1)

# Create a SpatialPointsDataFrame from rf_dat
coordinates(lamr_subset) <- ~longitude + latitude

# Assign the CRS from your raster (converted to PROJ.4 format)
proj4string(lamr_subset) <- crs(d2_lake, proj = TRUE)


#====================================================


#subset rf_dat to contain only important_covariates
rf_dat <- lamr_subset[, c("d2_lake", 
                          "d2_ancient", 
                          "spp3_cvr", 
                          "d2_aqua", 
                          "bedrock", 
                          "residuals")]

# Create experimental variogram of residuals
vgram <- variogram(residuals ~ 1, rf_dat)

#define good starting values
initial_model <- vgm(
  psill = var(rf_dat$residuals) / 2,
  model = "Sph",
  range = max(vgram$dist) / 3,
  nugget = var(rf_dat$residuals) / 4
)

# Fit variogram model
vgram_model <- fit.variogram(vgram, initial_model)

plot(vgram, vgram_model)

#compare multiple model types
models <- c("Sph","Exp","Gau","Mat")

fits <- lapply(models, function(m) {
  fit.variogram(vgram,
                vgm(psill = var(rf_dat$residuals)/2,
                    model = m,
                    range = max(vgram$dist)/3,
                    nugget = var(rf_dat$residuals)/4))
})

names(fits) <- models
fits 

#store the best, fitted variogram - Sph
library(gstat)

# Fit spherical variogram with your chosen parameters
vgram_model <- fit.variogram(vgram, model = vgm(psill = 0.161, model = "Sph", range = 8.857, nugget = 0.798))
plot(vgram, vgram_model)
```

Now that we have a variogram, we can begin building the regression kriging process

```{r}
# 1. stack the predictors into a single raster stack
d2_aqua    <- project(d2_aqua, d2_lake)

predictors <- rast(list(d2_lake, d2_ancient, spp3_cvr, d2_aqua))
names(predictors) <- c("d2_lake", "d2_ancient", "spp3_cvr", "d2_aqua")

# 2. create my model object
library(lme4)

# 3. convert spatial data to a regular data frame
lamr_df <- as.data.frame(lamr_subset)

# If you want coordinates as columns (optional)
lamr_df$lon <- coordinates(lamr_subset)[,1]
lamr_df$lat <- coordinates(lamr_subset)[,2]

# 4. Create a mixed-effects model object with watershed as random effect
model <- lmer(lake_amr ~ d2_lake + d2_ancient + spp3_cvr + d2_aqua + (1 | bedrock), REML = TRUE,
              data = lamr_df)
summary(model)

# 5. Create a spatial object for the residuals
res_sf <- st_as_sf(lamr_subset)

# 6. Predict fixed effects over the raster grid
#convert layers to numeric
predictors$d2_lake    <- as.numeric(predictors$d2_lake)
predictors$d2_ancient <- as.numeric(predictors$d2_ancient)
predictors$spp3_cvr   <- as.numeric(predictors$spp3_cvr)
predictors$d2_aqua    <- as.numeric(predictors$d2_aqua)

pred_df <- as.data.frame(predictors, xy = TRUE, NA.RM=TRUE)
pred_df$pred_fixed <- predict(model, newdata = pred_df, re.form = NA)  
# re.form = NA ensures only fixed effects are used in the prediction

# 7. Convert to SpatialPointsDataFrame for kriging residuals
coordinates(pred_df) <- ~ x + y

# 8. create my spatial data
coordinates(lamr_df) <- ~ lon + lat
proj4string(lamr_df) <- CRS("+proj=longlat +datum=WGS84")
#If not already set, do the following:
#coordinates(pred_df) <- ~ x + y  # x and y come from raster grid
proj4string(pred_df) <- CRS("+proj=longlat +datum=WGS84")

# 9. Krige residuals
res_krige <- krige(residuals ~ 1, lamr_df, pred_df, model = vgram_model)

# 10. combine residual kriging with mixed model predictions
# predict fixed effects at the same grid (without random effects)
pred_df$pred_fixed <- predict(model, newdata = as.data.frame(pred_df), re.form = NA)

# add kriged residuals
pred_df$rk_pred <- pred_df$pred_fixed + res_krige$var1.pred
```

Now that we have the final model predictions and the residuals kriged, we can rasterize our points and export to a geotiff. 
```{r}
# 1. Get min/max coordinates
xmin <- min(pred_df$x)
xmax <- max(pred_df$x)
ymin <- min(pred_df$y)
ymax <- max(pred_df$y)

# 2. Decide resolution
res_xy <- 0.01  # e.g., 0.01 degrees, adjust as needed

# 3. Compute number of rows and columns
ncol <- ceiling((xmax - xmin) / res_xy)
nrow <- ceiling((ymax - ymin) / res_xy)

# 4. Create raster template
r_template <- rast(nrows = nrow, ncols = ncol,
                   xmin = xmin, xmax = xmax,
                   ymin = ymin, ymax = ymax,
                   crs = "+proj=longlat +datum=WGS84 +no_defs")  # string CRS


# 5. rasterize
# Convert SpatialPointsDataFrame to SpatVector
pred_vect <- vect(pred_df)  # terra-friendly

# Then rasterize
AMR_reg_krige <- rasterize(pred_vect, r_template, field = "rk_pred")

# 6. Plot to make sure I get the variation I expect
library(RColorBrewer)

# Create a warm-to-cool palette: red = high, blue = low
pal <- colorRampPalette(c("lightblue", "dodgerblue", "orange", "red", "red4"))(100)

plot(AMR_reg_krige, main="AMR Predictions", col=pal)
plot(AMR_reg_krige, main="AMR Predictions", col=viridis::viridis(50))

# 6. Export as a GeoTIFF
writeRaster(AMR_reg_krige, "R_out/Htest_AMR_reg_krige.tif", overwrite = TRUE)

```

Spatial LOOCV

```{r}
###############################################################################
# Spatial Leave-One-Out Cross Validation for Regression Kriging
###############################################################################

library(gstat)
library(lme4)
library(sp)

# 1. Create the dataframe
lamr_df <- as.data.frame(lamr_subset)

# Ensure the coordinates are numeric
lamr_df$lon <- as.numeric(lamr_df$longitude)
lamr_df$lat <- as.numeric(lamr_df$latitude)
lamr_df$lake_amr <- lake_amr

# 2. PRE-SCALE variables safely
# Note: I am using the 'mean_' prefix names from your earlier subsetting
# If you renamed them to d2_lake etc. already, use those names instead.
lamr_df$d2_lake_s    <- as.numeric(lamr_df$d2_lake)
lamr_df$d2_ancient_s <- as.numeric(lamr_df$d2_ancient)
lamr_df$spp3_cvr_s   <- as.numeric(lamr_df$spp3_cvr)
lamr_df$d2_aqua_s    <- as.numeric(lamr_df$d2_aqua)

n_points <- nrow(lamr_df)
predictions <- numeric(n_points)

# Define Global FIXED Variogram Model
fixed_v_model <- vgm(psill = 0.161, model = "Sph", range = 8.857, nugget = 0.798)

for(i in 1:n_points){

  cat("Processing lake", i, "of", n_points, "\n")

  # Split data into training and testing frames
  train_df <- lamr_df[-i, ]
  test_df  <- lamr_df[i, ]

  # Create spatial training object for Kriging residuals
  train_sp <- train_df
  coordinates(train_sp) <- ~ lon + lat
  proj4string(train_sp) <- CRS("+proj=longlat +datum=WGS84")

  # 3. Fit regression model
  model_cv <- lmer(
    lake_amr ~ d2_lake_s + d2_ancient_s + 
               spp3_cvr_s + d2_aqua_s + (1 | bedrock),
    data = train_df
  )

  # 4. Calculate residuals
  train_sp$resid <- residuals(model_cv)

  # 5. Krige residual to the single test point
  # Create a spatial point for the test lake
  test_sp <- test_df
  coordinates(test_sp) <- ~ lon + lat
  proj4string(test_sp) <- CRS("+proj=longlat +datum=WGS84")
  
  kr <- krige(resid ~ 1, train_sp, test_sp, model = fixed_v_model)

  # 6. Predict fixed effects for the test point
  reg_pred <- predict(model_cv, newdata = test_df, re.form = NA)

  # 7. Combine
  predictions[i] <- reg_pred + kr$var1.pred
}

###############################################################################
# Evaluate model
###############################################################################
observed <- lamr_df$lake_amr
rmse <- sqrt(mean((predictions - observed)^2))
r2   <- cor(predictions, observed)^2
correlation <- cor(predictions, observed)

cat("LOOCV RMSE:", rmse, "\n")
cat("LOOCV R²:", r2, "\n")
cat("LOOCV correlation:", correlation, "\n")

lamr_df$LOOCV_pred <- predictions
```

Now we can make some diagnostic plots to show how well (or not) out model is working. 

```{r}
library(patchwork)

# --- Create the plot_df for Diagnostics ---
plot_df <- data.frame(
  observed = observed,
  predicted = predictions,
  residuals = observed - predictions,
  x = lamr_df$lon,
  y = lamr_df$lat
)

# Define your color scale (required for the p1, p2, p4 plots)
# This creates a consistent blue-to-red scale centered at 0
my_color_scale <- scale_color_gradient2(
  low = "blue", 
  mid = "white", 
  high = "red", 
  midpoint = 0, 
  name = "Residual"
)

# Make sure your Vancouver Island outline is ready for Panel D
# In your setup it was vi_vect, but Panel D calls vi_sf
vi_sf <- st_as_sf(vi_vect)

# Define a "Publication Theme" to keep everything consistent
# base_size = 14 or 16 is best for multi-panel figures
pub_theme <- theme_minimal(base_size = 15) + 
  theme(
    plot.title = element_text(size = 20, face = "bold"), # Bigger a, b, c, d
    axis.title = element_text(size = 16),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

# --- Panel A ---
# --- Panel A (Corrected) ---
p1 <- ggplot(plot_df, aes(x = observed, y = predicted, color = residuals)) +
  geom_point(alpha = 1, size = 3) + 
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", linewidth = 1.2) +
  my_color_scale +
  labs(x = "Observed AMR", y = "Predicted AMR", title = "a)") +
  pub_theme +
  # Use 'predictions' (the vector) instead of 'predicted' (the column name)
  annotate("text", x = min(observed) + 0.1, y = max(predictions),
           label = paste0("RMSE = ", round(rmse,2), "\nR = ", round(correlation,2)),
           hjust = 0, vjust = 2, size = 5.5)

# --- Panel B ---
p2 <- ggplot(plot_df, aes(x = predicted, y = residuals, color = residuals)) +
  geom_point(alpha = 1, size = 3) + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  my_color_scale +
  labs(x = "Predicted lake_amr", y = "Residuals", title = "b)") +
  pub_theme

# --- Panel C ---
p3 <- ggplot(plot_df, aes(x = residuals)) +
  geom_histogram(bins = 15, fill = "gray80", color = "black") + 
  geom_vline(xintercept = mean(residuals, na.rm = TRUE), color = "red", 
             linetype = "dashed", linewidth = 1.1) +
  labs(x = "Residuals", y = "Count", title = "c)") +
  pub_theme

# --- Panel D ---
p4 <- ggplot() +
  geom_sf(data = vi_sf, fill = NA, color = "black", linewidth = 0.6) +
  geom_point(data = plot_df, aes(x = x, y = y, color = residuals), size = 3.5) +
  my_color_scale +
  labs(title = "d)") +
  pub_theme +
  coord_sf()

# --- Final Assembly ---
final_plot <- (p1 + p2) / (p3 + p4) + 
  plot_layout(guides = "collect") & 
  theme(legend.position = "right")

# Note: ggsave is best for keeping text crisp in publications
ggsave("Final Models/LOOCV_Publication_Final.png", final_plot, width = 12, height = 10, dpi = 600)
```

